Cooperative Kernels: Safe Multitasking for Blocking Algorithms on GPUs

### 2016-11-15 write-ups/PLDI/paper.tex
Trimming: losing some paragraph headings.

\paragraph{Patterns for irregular algorithms}
In \mysec\ref{sec:portingalgorithms} we describe the set of irregular GPU algorithms used
in our experiments, which largely captures the irregular blocking
algorithms that are available as open source GPU kernels.  These all
employ either work stealing or operate on graph data structures, and placing our new constructs follows a common, easy-to-follow pattern in each case.

Like the simplified example of \myfig\ref{fig:workstealing-cooperative}, the work stealing algorithms have a transactional flavour
and require little or no state to be carried between transactions.  Thus the point at which a workgroup is ready to process a new task is a natural point to place $\offerkill$ and $\offerfork$, and few or no $\transmit$ annotations are required.

The example of \myfig\ref{fig:cgraphsearch} is simpler than, but representative of,
most level-by-level graph algorithms.
It is typically the case that on completing a level of
the graph algorithm, the next level could be processed by more or
fewer workgroups, something which $\resizingglobalbarrier$
facilitates.  As in \myfig\ref{fig:cgraphsearch}, some level-specific state must be transmitted to new workgroups.

Our practical experience suggests that cooperative kernels are a good
fit for at least these two classes of irregular algorithms, which is
of course partly because we were inspired by these algorithms in
designing the model.

@@

\paragraph{Sufficient $\offerkill$ calls}

The purpose of $\offerkill$ is to provide the scheduler with an
opportunity to destroy a workgroup in order to schedule
higher-priority tasks.  The scheduler relies on the cooperative kernel
to execute $\offerkill$ sufficiently frequently that soft real-time
constraints of other workloads can be met.
%
Using our work stealing example: in
\myfig\ref{fig:workstealing-cooperative} a workgroup offers itself to
the scheduler after processing each task.  If tasks are sufficiently
fast to process then the scheduler will have ample opportunities to
de-schedule workgroups.  But if tasks can be very time-consuming to
process then it might be necessary to rewrite the algorithm so that
tasks are shorter and more numerous, to achieve a higher rate of calls
to $\offerkill$.

Getting this non-functional requirement right depends on the
constraints of the tasks to be co-scheduled with a cooperative kernel,
on properties of the cooperative kernel itself, and on the efficiency
of the GPU that is used for execution.  In \mysec\ref{sec:sizingnoncoop} we conduct
experiments to understand the response rate that would be required to
co-schedule graphics rendering with a cooperative kernel in order to
achieve a smooth frame rate, and we evaluate responsiveness in
practice in \mysec\ref{sec:responsiveness}.

\paragraph{Sufficient workgroups for the cooperative kernel}

Recall that the cooperative kernel would ideally utilise the $N$
workgroups requested upon kernel launch, in order to conduct its
processing efficiently.  The scheduler should thus aim to provide the
cooperative kernel $N$ workgroups if other constraints allow it,
by accepting an $\offerkill$ only if a compute unit is required for another
task, and responding positively to $\offerfork$ calls if compute units are available.  Our
implementation of cooperative scheduling (\mysec\ref{sec:implementation}) is as
generous as possible to the cooperative kernel in this regard.

Note that the above guidelines are not hard-and-fast rules.  For
example, a smart scheduler might respond selectively to $\offerkill$
and $\offerfork$ calls by predicting future high-priority workloads
based on past workloads.

@@

A goal of coopeative kernels is to allow traditional, non-blocking GPU
tasks (e.g.\ graphics shaders in OpenGL, Vulkan and DirectX, and regular data-parallel computational kernels)
to be co-scheduled with blocking irregular algorithms.  It is
important to note that our new model has no effect on the manner in
which these traditional GPU workloads are programmed.

Furthermore, if our cooperative language extensions are
ignored---$\offerkill$ and $\offerfork$ are treated as no-ops,
$\transmit$ annotations are ignored, and $\globalbarrier$ and
$\resizingglobalbarrier$ calls are redirected to an existing global
barrier implementation (so that no resizing occurs in the latter
case)---then a cooperative kernel will behave just like its
non-cooperative counterpart would behave.  As a result, a developer
for whom today's occupancy-bound execution model suffices can upgrade
their kernel to a cooperative form and retain backwards-compatibility
with today's GPU platforms by pre-processing away our language
extensions.

>>>>

\paragraph{Patterns for irregular algorithms}
In \mysec\ref{sec:portingalgorithms} we describe the set of irregular GPU algorithms used
in our experiments, which largely captures the irregular blocking
algorithms that are available as open source GPU kernels.  These all
employ either work stealing or operate on graph data structures, and placing our new constructs follows a common, easy-to-follow pattern in each case.
%
The work stealing algorithms have a transactional flavour
and require little or no state to be carried between transactions.  The point at which a workgroup is ready to process a new task is a natural place for $\offerkill$ and $\offerfork$, and few or no $\transmit$ annotations are required.
%
\myfiglong\ref{fig:cgraphsearch} is simpler than, but representative of,
most level-by-level graph algorithms.
It is typically the case that on completing a level of
the graph algorithm, the next level could be processed by more or
fewer workgroups, which $\resizingglobalbarrier$
facilitates.  Some level-specific state must be transmitted to new workgroups.

Our experience is that cooperative kernels are a good
fit for at least these two classes of irregular algorithms, in part because they inspired the model.

@@

%\paragraph{Sufficient $\offerkill$ calls}

The purpose of $\offerkill$ is to provide the scheduler with an
opportunity to destroy a workgroup in order to schedule
higher-priority tasks.  The scheduler relies on the cooperative kernel
to execute $\offerkill$ sufficiently frequently that soft real-time
constraints of other workloads can be met.
%
Using our work stealing example: a workgroup offers itself to
the scheduler after processing each task.  If tasks are sufficiently
fast to process then the scheduler will have ample opportunities to
de-schedule workgroups.  But if tasks can be very time-consuming to
process then it might be necessary to rewrite the algorithm so that
tasks are shorter and more numerous, to achieve a higher rate of calls
to $\offerkill$.
%
Getting this non-functional requirement right is GPU- and application-dependent.
In \mysec\ref{sec:sizingnoncoop} we conduct
experiments to understand the response rate that would be required to
co-schedule graphics rendering with a cooperative kernel, maintaining a smooth frame rate.

%\paragraph{Sufficient workgroups for the cooperative kernel}

Recall that, on launch, the cooperative kernel requests $N$ workgroups.
The scheduler should thus aim to provide $N$ workgroups if other constraints allow it,
by accepting an $\offerkill$ only if a compute unit is required for another
task, and responding positively to $\offerfork$ calls if compute units are available.  Our
implementation (\mysec\ref{sec:implementation}) is as
generous as possible to the cooperative kernel in this regard.

These guidelines are not hard-and-fast rules, and e.g.\ a smart scheduler might respond selectively to $\offerkill$
and $\offerfork$ calls by predicting future high-priority workloads
based on past workloads.

@@

Our new model has no effect on the manner in which traditional GPU
workloads are programmed, such as graphics shaders in OpenGL, Vulkan
and DirectX, and regular data-parallel computational kernels.
%
Furthermore, if our cooperative language extensions are
ignored---$\offerkill$ and $\offerfork$ are treated as no-ops,
$\transmit$ annotations are ignored, and $\globalbarrier$ and
$\resizingglobalbarrier$ calls are redirected to an existing global
barrier implementation (so that no resizing occurs in the latter
case)---then a cooperative kernel will behave just like its
non-cooperative counterpart would behave.  A developer
for whom today's occupancy-bound execution model suffices can upgrade
their kernel to a cooperative form and retain backwards-compatibility
with today's GPU platforms by pre-processing away our language
extensions.
